---
title: Recapture rates and habitat associations of White-faced Darter _Leucorhinnia
  dubia_ on Fenn's and Whixall Moss, Shropshire, UK
author: "Rachel Davies, Achaz von Hardenberg & Matthew Geary"
date: "24 May 2018"
output:
  pdf_document: default
  html_document: default
---
## Abstract

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

## Introduction

There has been a marked decline in global biodiversity in the last several decades, a decline which is expected to continue, and this has been largely attributed to changes in land-use activities (REF - @sala_global_2000).Land-use activities include agriculture, forestry, creation of urban areas, and use of natural resources (REF - @foley_global_2005). These activities have a huge impact on environmental characteristics  and often cause habitat loss and fragmentation, contributing largely to the decline in global species diversity  (REF - @holloway_conervation_2003). As such management and protection of habitats and populations is vital at both a local and global scale (REF - @foley_global_2005; @holloway_conervation_2003). 

A bias exists in conservation research towards charismatic vertebrates (REF - @di_marco_changing_2017). Although Odonata are charismatic invertebrates they are not immune to this bias (REF - @clausnitzer_odonata_2009). In addition much research into Odonata focuses on physiology, evolution and behaviour (REF - @cordoba-aguilar_dragonflies_2008) and they have rarely been the focus of conservation research (REF - @clausnitzer_odonata_2009). Basic ecological research into demography, survival and habitat use is essential for effective protection of species and habitats. For any taxa this requires detailed ecological and life history data collected in the field. These are often difficult to obtain, particularly on large scales. Integrating large scale data such as presence-only distribution datasets with more detailed local information is a current challenge in conservation ecology (REF - @powney_beyond_2015).  

Methods to analyse habitat preferences are varied depending on the data available. The current 'gold standard' is the use of site occupancy models which take into account detectability (i.e. the probability that a species is detected in a site if present) when estimating occupancy: the probability that a species is present in a site (REF - @ackenzie_estimating_2003). Models using this framework help us to avoid the age-old problem of "imperfect detection", i.e. failing to spot a species during a survey on a site where it is actually present; REF - @ackenzie_estimating_2003). However, these models require repeated surveys where both detections and non-detections are recorded; these data are not always available, especially in the case of historic data collected by volunteers. On larger scales a number of methods exist which can use only presence records along with environmental covariates (REF - @elith_species_2009). These can tell us about habitat use but are constrained to estimate a measure of the relative importance of habitats rather than the true probability of presence (REF - @lith_statistical_2011) and are limited by the environmental data available. At very small scales, such as individual protected areas, detailed data on habitats and landcover can be difficult to obtain. Datasets such as the UK lancover map (LCM2015), although the resolution is 15m, are too crude for local studies in some areas. Simpler methods which indicate preferred habitat, such as selection indices (REF - @manly_resource_2007), have fewer assumptions and can be revealing even at small scales (REF - @neu_technique_1974).

Investigating survival and movement requires individual recognition and methods using a capture mark-recapture approach are well established (REF- @mccrea_analysis_2014). Such analyses can tell us about the age-sex specific survival probabilites of individuals, the use of different sites or habitats and how these changes over time and the likelihood of encountering individuals again in the future. High quality data of this type can provide accurate estimates of population size. Mark-recapture methods have been used on Odonata populations in the past to monitor rare species as well as to study the  life history of more abundant species (REF - @oster_evaluating_2004; @cordero-rivera_mark-recapture_2008) to monitor rare species as well as to study the life history of more abundant species (REF - @anholt_mark-recapture_2001). Odonata, because of the ease by which they can usually be individually marked, have also been used as  model species for methodological research on the development of mark-recapture techniques (REF - @manly_b.f.j._new_1968). 

The White-faced Darter (_Leucorrhinia dubia_), is a specialist of lowland peatbogs where it breeds in bog 
pools containing sphagnum mosses (REF - @smallshire_britains_2004). It has a life cycle that includes a 1-3-year larval period, followed by an adult flight period (REF - @smallshire_britains_2004). Emergence is weather dependent and will typically start in either May or June each year. Tenerals are thought to disperse to low scrub following emergence, staying there whilst they mature. Following this, the
adults return to breeding pools, with males returning sooner than females so they can hold breeding territories (REF - @smallshire_britains_2004). The adult flight period typically ends in either late July or August. The White-faced Darter has a scattered distribution and its populations have been declining in Britain over the past several decades. Despite being classified as a species of least concern on the IUCN Red Data List (REF - @clausnitzer_odonata_2009), this decline in Britain has resulted in a classification of Endangered on the Odonata Red Data List for Britain (REF - @beynon_odonata_2008). This decline is largely attributed to habitat loss and the resulting habitat fragmentation (REF - @beynon_odonata_2008); as over 90% of England’s peatbogs have been lost or substantially damaged to date (REF - @nature_peat_2002). There are currently only three stable historical populations of White-faced Darter in England, along with two recently reintroduced populations, one in Cumbria and one in Cheshire (REF - @clarke_white-faced_2014; @meredith_reintroduction_2017).

Here we use two methods to investigate important ecological characteristics of White-faced Darter on Fenn's and Whixall moss in Shropshire, UK. We use mark-recapture methods to investigate survival and movements of adults during the flight period and a selection index method to investigate habitat use. These methods can both contribute to our understanding of the spatial use of habitat by White-faced Darter and can help us to prioritise future research for this species. 


## Methods

### Study area  
Fenn’s, Whixall and Bettisfield Mosses (FWB Mosses) are located within Shropshire (52°55′N 2°46′W) and they support a large, long-established population of White-faced Darter. FWB Mosses are a lowland raised bog complex, stretching nearly 1000 hectares (REF - @meredith_reintroduction_2017). Historically, the mosses were used for peat cutting and in the 19 th century they were drained to allow larger-scale operations to take place (REF - @meredith_reintroduction_2017). Eventually, in 1990, the mosses were taken over by English Nature (now Natural England) and long-term restoration began, benefitting a whole host of mossland species, including the White-faced Darter. (REF - @meredith_reintroduction_2017).

### Field methods

The site was surveyed twice per week between the 22nd of May and 6th of July 2017. This encompassed the peak flight period of White-faced Darter (REF - @smallshire_britains_2004). Two separate breeding pools within FWB Mosses were sampled simultaneously, along with a variety of scrub and other potentially suitable habitat. On each sampling occasion, the full sampling area was searched for any White-faced Darter individuals. Different routes were walked on each occasion to allow different areas within the sampling area to be searched at different times of the day. Sampling sessions lasting between 5-10 hours, being carried out between 10am and 4pm, as this is the favoured flight period for adult dragonflies (REF - @smallshire_dragonfly_2010). Sampling days were weather dependent (REF - @chin_interactive_2009) and weather conditions were recorded on all sampling days. 


_Capture-Mark-Recapture_  
Where possible, mature adults were caught using an invertebrate net and marked with a unique number on their wing (REF - @chin_interactive_2009), using an Edding 404 permanent marker pen (Plate.1 - NEED TO ADD IN PICTURE - IS THAT OK RACHEL?). The insects were then released at point of capture and any behavioural observations recorded. Not all observed individuals were captured and tenerals were excluded from the capture-mark-recapture survey as during this life stage they are fragile and handling may cause wing damage (REF - @allen_population_2014). Tenerals are easily identified by their pale green colouration, a lack of their full adult colouration and by their shiny wings (REF - @smallshire_britains_2004). Insects recaptured on day of marking were not re-counted (REF - @foster_evaluating_2004). Following an initial marking, recapture on successive days was only necessary when relevant information could not be collected from re-sighting individuals (REF - @lettink_introduction_2003).

_Selection index_
White-faced Darter presence was recorded while searching the site during the capture-mark-recapture study. This included captured individuals as well as those seen on survey routes but not captured. On each occasion the location of the individual was recorded with a hand-held GPS unit (Garmin GPSMAP 64). Additionally, a phase 1 habitat survey (REF - @committee_handbook_2010) was conducted across the study site to produce a habitat map using 100 x 100 m grid cells. The proportion of five habitat types were recorded in each square: moss (peat moss, rushes and sedges), scrub (low woody vegetation), scrub-moss (peat moss with low woody vegetation), water (open pools) and woodland (mature trees). From this the dominat habitat in each square was calculated. Of these, only water was not used in analyses as adult individuals tended to be sighted over terrestrial habitat. 

### Data analysis

_Capture-mark-recapture_  
A single-season Cormack-Jolly-Seber model (REF - @pradel_utilization_1996) was used to determine to probability of survival between sampling days and the probability of recapture on sampling days for the capture-mark-recapture data. Capture-mark-recapture analysis was carried out using the 'RMark' package version 2.2.2 (REF - @fouchet_r_2015), in R version 3.3.2. (REF - @r_core_team_r:_2016 <!--- Is this the right version now? Doubt it... --->).

```{R capture_mark_recapture_analyis}


library(RMark)
#require(reshape)

################
#function to read the matrix and create the capture history strings
pasty<-function(x)
  {
    k<-ncol(x)
    n<-nrow(x)
    out<-array(dim=n)
    for (i in 1:n)
      {
        y<-(x[i,]>0)*1
        out[i]<-paste(y[1],y[2],y[3],y[4],y[5],sep="")
        }
    return(out)
    }

##################

input.data<-read.csv("Book3.csv")
#head(input.data)

read.csv("now.csv")
now<-read.csv("now.csv")
x <- cbind(now$Week,now$Week)
y <- cbind(now$Male,now$Female)
matplot(x,y,type="p", pch = 16, bty = "n", xlab = "Week of the Year", ylab = "Number of Individuals Marked")
legend("topright", legend=c("Male", "Female"), pch=16, col=c("black","red"))


#capture history data frame for RMark
capt.hist <- data.frame(ch=pasty(input.data[,3:14]))
#View(capt.hist)
capt.hist1 <- data.frame(ch=pasty(input.data[,3:14]), sex = input.data$sex)
#View(capt.hist1)
capt.hist1 <- capt.hist1[capt.hist1$ch != '00000',]

#process data to CJS format
#necessary because sex is a factor (group) not a continuous covariate
capthist1.processed=process.data(capt.hist1,model="CJS",groups="sex")

#this is not necessary because it is automatically created in the mark() function.
#but it allows us to see (and change) all the details of the design data
caphis.ddl<-make.design.data(capthist1.processed)

###run.models(model.list= NULL, type= "CJS")
### says no models need to be run?

Phi.dot_p.dot=mark(capthist1.processed)
# This output looks like it has been confused by sex

# Writing a new model that specifies interactive sex and time effects

Phi.sex.T=list(formula=~sex*time)
p.sex.T=list(formula=~sex*time)
Phi.sex.T_p.sex.T=mark(capthist1.processed,model.parameters=list(Phi=Phi.sex.T,p=p.sex.T))
##still looks like the model is being confused by sex


Phi.sex=list(formula=~sex-1)
p.sex=list(formula=~sex-1)
Phi.sex_p.sex=mark(capthist1.processed,model.parameters=list(Phi=Phi.sex,p=p.sex))
##still looks like the mo

Phi.sex=list(formula=~1)
p.sex=list(formula=~sex)
Phi.1_p.sex=mark(capthist1.processed,model.parameters=list(Phi=Phi.sex,p=p.sex))
##still looks like the mo


Phi.t=list(formula=~time)
p.1=list(formula=~1)
Phi.t_p.1=mark(capthist1.processed,model.parameters=list(Phi=Phi.t,p=p.1))


#This is the model that gives us the result we need
Phi.dot_p.dot=mark(capthist1.processed)
```

_Selection index_  
Selection indices calculate habitat use as a ratio between habitat where a species is recorded compared to the proportion of each habitat within the study area (REF - @manly_resource_200@7). Although relatively simple they can be effective in indicating habitat use (REF - @manly_resource_2007). Selection indices can be sensitive to the scale used in calculating habitat use however Neu's index is relatively robust to changes in scale (REF - @neu_technique_1974). For this reason we used Neu's index which calculates $w_{i} = \frac{u_{i}}{\pi_{i}}$ where $w_{i}$ is the proportion of squares of each dominant habitat type among all of the squares wiht White-faced Darter records and $\pi_{i}$ is the proportion of each dominat habitat type among all of the squares in the study area. Values of thsi index $> 1$ indicate use of a habitat type in greater proportions than it is generally available in the study area. Selection index analysis was performed in R version 3.4.4 (REF - @r_core_team_r:_2018).

```{r selection_index_analysis}

library(raster)
library(reshape2)
library(unmarked)


latlong <- "+init=epsg:4326"
ukgrid <- "+init=epsg:27700"


# Load presence data
### Wind score of 6+ changed to 7 for ease of interpretation
WFD.pres <- read.csv("FW_presdata.csv")
#plot(WFD.pres$Long, WFD.pres$Lat)

## Add 'presence' column
WFD.pres$pres <- 1

## Create spatial object for presence points
WFD.pres.latlong <- SpatialPointsDataFrame(data.frame(WFD.pres$Long, WFD.pres$Lat), data = WFD.pres, proj4string = CRS(latlong))

## Convert to BNG
WFD.pres.BNG <-  spTransform(WFD.pres.latlong,  CRS(ukgrid))
## Add trasnsformed x and y values to data frame
WFD.pres.BNG@data$x <- WFD.pres.BNG@coords[,1]
WFD.pres.BNG@data$y <- WFD.pres.BNG@coords[,2]


# Create habitat rasters
## Define extent
#fenn.ext <- extent(c(348000, 350000, 335000, 337000)) ## Previous extent needs correcting
fenn.ext <- extent(c(347890, 349890, 335000, 337000)) ## Corrected visually using Google maps

moss <- raster(as.matrix(read.csv("FW_hab/Fenns_Moss.csv", header = FALSE)), crs = crs(ukgrid))
extent(moss) <- fenn.ext               
writeRaster(moss, "FW_hab/Fenns_moss.img", overwrite = TRUE)

scrub <- raster(as.matrix(read.csv("FW_hab/Fenns_Scrub.csv", header = FALSE)), crs = crs(ukgrid))
extent(scrub) <- fenn.ext               
writeRaster(scrub, "FW_hab/Fenns_scrub.img", overwrite = TRUE)

scrubMoss <- raster(as.matrix(read.csv("FW_hab/Fenns_ScrubMoss.csv", header = FALSE)), crs = crs(ukgrid))
extent(scrubMoss) <- fenn.ext               
writeRaster(scrubMoss, "FW_hab/Fenns_scrubmoss.img", overwrite = TRUE)

water <- raster(as.matrix(read.csv("FW_hab/Fenns_Water.csv", header = FALSE)), crs = crs(ukgrid)) ## Water matrix corrected visually using aerial photographs (Google maps)
extent(water) <- fenn.ext               
writeRaster(water, "FW_hab/Fenns_water.img", overwrite = TRUE)

wood <- raster(as.matrix(read.csv("FW_hab/Fenns_Wood.csv", header = FALSE)), crs = crs(ukgrid))
extent(wood) <- fenn.ext               
writeRaster(wood, "FW_hab/Fenns_wood.img", overwrite = TRUE)

hab.dat <- data.frame(moss = getValues(moss), scrub = getValues(scrub), scrubMoss = getValues(scrubMoss), water = getValues(water), wood = getValues(wood))

### Funtion to select and print dominant habitat
## x is data frame of habitat proportions
## n is names of columns
dom.hab <- function(x, n){
   if(is.na(max(x))){
     y <- NA
   }
  else{
   y <- n[which.max(x)]
  }
    return(y)
}
hab.names <- names(hab.dat)
dom.hab.vals <- character(400)
for(i in 1:length(hab.dat[,1])){
  dom.hab.vals[i] <- dom.hab(hab.dat[i,], hab.names)
}
hab.dat$dom <- as.factor(dom.hab.vals)

pres.hab <- data.frame(moss = extract(moss, WFD.pres.BNG), scrub = extract(scrub, WFD.pres.BNG), scrubMoss = extract(scrubMoss, WFD.pres.BNG), water = extract(water, WFD.pres.BNG), wood = extract(wood, WFD.pres.BNG))
pres.hab.vals <- character(258)
for(i in 1:length(pres.hab[,1])){
  pres.hab.vals[i] <- dom.hab(pres.hab[i,], hab.names)
}
pres.hab$dom <- as.factor(pres.hab.vals)

w.i <- table(pres.hab$dom)/sum(table(pres.hab$dom))
pi.i <- table(hab.dat$dom)[-4]/sum(table(hab.dat$dom)[-4])


selection.index <- as.vector(w.i)/as.vector(pi.i)
names(selection.index) <- c("Open moss", "Scrub", "Scrub & Moss", "Woodland")
```



## Results

_Capture-Mark-Recapture Model_  
A total of 13 sampling days were carried out at FWB Mosses from the 22nd May 2017 until the 7th July 2017. During these sampling days, a total of 50 adult White-faced Darter were marked (Figure!!!) (41 males, 9 females), and a total of 6 recaptures were made. Probability of survival between sampling days was estimated at 0.25 (SE = 0.67, 95% confidence intervals: 0.08, 0.57). Probability of capture on each sampling day was estimated at 0.39 (SE = 0.76, 95% confidence intervals: 0.06, 0.86). Further models using a range of co-variates were unsuitable as the models became over parameterised due to the lack of recaptures. 

<!--Why dont we use the CMR data to estimate also population size in the site? // Yes, let's do this-->
```{r capture_mark_recapture_figure_1}
mark<-read.csv("mark.csv")
plot(mark$Number ~ mark$Sex)
fen<-read.csv("fen.csv")

plot(fen$Week, fen$Total, type = "o", pch = 16, xlab = "Week of the Year", ylab= "Number of Individuals Recorded")
```

```{r capture_mark_recapture_figure_2}
plot(fen$Week, fen$Total, xlab= "Week of the Year", ylab = "Total Number of L.dubia Individuals Recorded")
```

```{r capture_mark_recapture_figure_3}
df <- read.csv("fen.csv",sep=",",head=T)
x <- cbind(df$Week,df$Week,df$Week)
y <- cbind(df$Male,df$Female,df$Teneral)
matplot(x,y,type="p", pch = 16, bty = "n", xlab = "Week of the Year", ylab = "Number of Individuals Recorded")
legend("topright", legend=c("Male", "Female", "Teneral"), pch=16, col=c("black","red","green"))
```

_Selection Index_  
A further 304 individual White-faced Darter were observed during the fieldwork, 234 of which were not captured but only observed from distance (Figure!!!) <!--again, this last sentence is more a methodological issue rather than a result, no? // I agree but I'm also bulking out this paragraph in the results section, let me off just this once?-->. White-faced Darter show a clear preference (SI > 1) for 'moss' habitats while scrub and woodland (smallest SI) appear to be avoided (Figure!!!). <!-- we could get confidence intervals around these estimates by bootstrapping maybe-->

```{r selection_index_figure}
barplot(selection.index, ylab = "Neu's selection index", xlab = "Habitat type")
#Neu's selection index for White-faced Darter records on Fenn's and Whixall Moss
```

### Discussion

The capture-mark-recapture model suggested that the survival rate of adult White-faced Darter from one survey session to the next was 25%, however, the confidence intervals around this estimate were very wide. Similarly, we estimated a 40% chance of each individual being recaptured by our methods but with very wide confidence intervals. The reason for the wide confidence intervals around these estimates was the very low recapture rate (only 6 recaptures in 13 survey days). Although low capture rates might be expected in a large invertebrate population and have been noted before in Odonata (REF - @cordero-rivera_mark-recapture_2008), this was lower than expected. Although male White-faced Darter hold territories they are less tied to these sites than species such as Four-spotted chaser (_Libellula quadrimaculata_) and so are less predictable in their movements (REF - @merritt1996atlas). We suggest that in future capture-mark-recapture approaches for this species, and other similarly cryptic species, need a greater number of capture days and more researchers in the field making captures. This increase in effort is likely to increase the capture rate and increase the accuracy of estimates. 

Many more White-faced Darter were seen than were captured and the resource selection index calculated using these data suggest that they prefer the 'moss' habitat among those available. Although this habitat is the most common habitat in the study area, the selection index suggests that they use this habitat in proportions greater than those available across the site. The 'moss' habitat consists of peat with low heather vegetation and wet flushes and is the habitat most commonly found at pool edges. This is the habitat described in previous research on White-faced Darter (REF - @dolny_how_2018) and described in Boudot & Kalman (REF - @boudot_atlas_2015) including "peat moss, rushes and sedges". Locally on this site, White-faced Darter appear to avoid complex vegetation, including scrub and woodland. However, White-faced Darter sites, especially those in Scotland which represents the stronghold for this species in Britain, are often forested (REF  - @cham2014atlas). Breeding pools within these sites are likely to be in open areas but the association with woodland, particularly ancient woodland (REF - @cham2014atlas), is suggestive of some associations between White-faced Darter and these habitats at larger scales.

Moss habitats are certainly suitable for White-faced Darter, however, the low capture and recapture rate we found in this study may explain why open 'moss' habitats appeared to be preferred for this species. White-faced Darter are well camouflaged within their habitats and, as such, there is a good chance of missing individuals because of habitat complexity (i.e. low detectability, REF - @mazerolle_making_2007). Unfortunately, our field methods, did not allow us to estimate detectability in terms of sightings on this occasion but the low capture probability found in our capture-mark-recapture study suggests it is very low.  In future we suggest that survey methods are designed so that detectability can be estimated explicitly, in order to get more accurate estimates of occupancy and thus of resource selection. In this case we are left unable to confidently suggest whether White-faced Darter are avoiding more complex vegetation or whether they are harder to see and therefore record in these habitats. 

Data which allows the estimation of detectability can easily be collected with just a few minor changes to currently common survey methods. In fact, the majority of these suggestions are already being requested by the BDS to provide data for the upcoming State of the Nation's Dragonflies in 2020. We would like to add our voice to these calls to record complete lists and to repeat site visits. Complete lists are records of all the Odonata species detected on a single visit and allow non-detection to be inferred where species are not recorded (REF - @isaac_bias_2015). This requires recorders to note very common species as well as rarities. Unfortunatley, there is a tendency in biological recordings to note only the rare or exciting species (e.g. first record of the year) and this can bias our inferences about population change amongst more common species (REF - @isaac_statistics_2014). Repeated site visits allow us to estimate the detectability of a species (REF - mackenzie_estimating_2003) and consequently get unbiased estimates of occupancy, not affected by imperfect detection. We would also like to suggest that where possible recorders include some measure of effort in their surveys (e.g. time spent surveying or distance walked). Ideally this would be standardized and included in official protocols such as those already commonly in use for bird surveys. However, this suggestion is less vital because effort could potentially be crudely estimated through list-length analysis (REF- @szabo_regional_2010) provided that multiple visits are made to the same sites <!--I have some doubts about this tbh: particular good sites will provide longer lists, regardless of effort, no? // It's a relative measure which is why repeated visits are needed. We could get rid of this sentence if you want? -->.

We present the results in this paper as an indication of what can be done in terms of conservation research in Odonata. Although we have been unable to make firm inferences regarding White-faced Darter survival and habitat preference at this stage this study can provide valuable information which can contribute to the design of future studies. We suggest that research into the conservation ecology of White-faced Darter along with other Odonata species threatened with declining ranges, declining populations or habitat loss is essential to the long-term conservation of these species. Methods for such studies can be well informed by current practices used with other taxa. In particular, the analytical advances made in ornithology, research on Lepidoptera and work related to the use of data collected through citizen science provide a fantastic opportunity to advance our knowledge on the conservation ecology of Odonata. 

### Acknowledgements

We would like to express our thanks to the British Dragonfly Society who supported this work financially, Natural England who provided permission to access the field site and to Chris Meredith who provided invaluable help in the field and with advice on White-faced Darter ecology and behaviour. We would also like to thank all of the volunteers who helped with fieldwork in the summer of 2017.

### References

